---
title: "Screening Task Notebook_TASK_1"
author: "TRAN NHAT VAN NHU"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r}
# Code from Task_1.R
# Load libraries 
library(Seurat)
library(SeuratObject)
library(hdf5r)
library(ggplot2)
library(viridis)
library(png)
library(grid)
library(glmGamPoi)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(stringr)
library(ggpubr)
traceback()

#----------------------------------------------
# Step 1: Define base directory and sample IDs
#----------------------------------------------
base_dir <- "C:/Users/ADMIN/Desktop/GSE245620_RAW"
sample_ids <- c("PSC011_A1","PSC011_B1","PSC011_C1","PSC011_D1")
```



```{r}
#-------------------------------------------------
# Step 2: Function to process counts + coordinates
#-------------------------------------------------
process_spatial <- function(base_dir, sample_id) {
  sample_dir <- file.path(base_dir, sample_id)
  
  # 2.1 Read counts
  counts <- Read10X(data.dir = sample_dir)
  obj <- CreateSeuratObject(counts = counts, assay = "Spatial")
  
  # 2.2 Read spatial coordinates
  coords <- read.csv(file.path(sample_dir, "spatial/tissue_positions.csv"), header = FALSE)
  colnames(coords) <- c("barcode","in_tissue","array_row","array_col","pxl_row","pxl_col")
  obj@meta.data <- cbind(obj@meta.data, coords[match(rownames(obj@meta.data), coords$barcode), ])
  
  return(obj)
}

# Run for all replicates
PSC011_A1 <- process_spatial(base_dir, "PSC011_A1")
PSC011_B1 <- process_spatial(base_dir, "PSC011_B1")
PSC011_C1 <- process_spatial(base_dir, "PSC011_C1")
PSC011_D1 <- process_spatial(base_dir, "PSC011_D1")

PSC011_A1$sample <- "sample_A1"
PSC011_B1$sample <- "sample_B1"
PSC011_C1$sample <- "sample_C1"
PSC011_D1$sample <- "sample_D1"

```



```{r}
#--------------------------------------------
# Step 3: Process integrated spatial datasets
#--------------------------------------------
PSC011_all <- merge(
  PSC011_A1, 
  y = list(PSC011_B1, PSC011_C1, PSC011_D1),
  add.cell.ids = c("A1","B1","C1","D1"),
  project = "PSC011")
PSC011_all <- JoinLayers(PSC011_all)

# Standard preprocessing
PSC011_all[["percent.mt"]] <- PercentageFeatureSet(PSC011_all, pattern = "^MT-")   #mitochondrial percentage
VlnPlot(PSC011_all,
  features = c("nFeature_Spatial","nCount_Spatial","percent.mt"), ncol = 3)

PSC011_all <- subset(
  PSC011_all,
  subset =
    nFeature_Spatial > 500 &
    nFeature_Spatial < 5000 &
    percent.mt < 15)

PSC011_all <- NormalizeData(PSC011_all)     #Normalization
PSC011_all <- FindVariableFeatures(      # Find variable genes
  PSC011_all,
  selection.method = "vst",
  nfeatures = 3000)                       # best choice for spatial transcriptomics 2000–3000 genes

VariableFeaturePlot(PSC011_all)
top20 <- head(VariableFeatures(PSC011_all), 20)
LabelPoints(
  plot = VariableFeaturePlot(PSC011_all),
  points = top20,
  repel = TRUE)

# Spatial analysis pipeline
options(future.globals.maxSize = 10 * 1024^3)  # EXTEND CAPACITY 10 GB
PSC011_all <- ScaleData(PSC011_all)
PSC011_all <- RunPCA(PSC011_all)
ElbowPlot(PSC011_all) #dims = 1:20
PSC011_all <- FindNeighbors(PSC011_all, dims = 1:20)
PSC011_all <- FindClusters(PSC011_all, resolution = 0.5)
PSC011_all <- RunUMAP(PSC011_all, dims = 1:20)

DimPlot(PSC011_all, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
DimPlot(PSC011_all, group.by = "sample")

# Load H&E image separately
process_image <- function(base_dir, sample_id) {
  sample_dir <- file.path(base_dir, sample_id)
  img <- readPNG(file.path(sample_dir, "spatial/tissue_hires_image.png"))
  g <- rasterGrob(img, width = unit(1,"npc"), height = unit(1,"npc"))
  
  return(g)
}

# Run image loading for all replicates
PSC011_A1_img <- process_image(base_dir, "PSC011_A1")
PSC011_B1_img <- process_image(base_dir, "PSC011_B1")
PSC011_C1_img <- process_image(base_dir, "PSC011_C1")
PSC011_D1_img <- process_image(base_dir, "PSC011_D1")
```



```{r}
#----------------------------------------------
# Step 4: Quality control on spatial spots
#----------------------------------------------
# total counts per spot
ggplot(PSC011_all@meta.data, aes(x = pxl_col, y = pxl_row, color = nCount_Spatial)) +
  geom_point(size = 3.0) +
  scale_color_viridis_c() +
  #scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  ggtitle("PSC Liver Tissue: Count per spot")

# Number of features per spot
ggplot(PSC011_all@meta.data, aes(x = pxl_col, y = pxl_row, color = nFeature_Spatial)) +
  geom_point(size = 3.0) +
  scale_color_viridis_c() +
  #scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  ggtitle("PSC Liver Tissue: Feature per spot")

ggplot(PSC011_all@meta.data, aes(x = nCount_Spatial)) +
  geom_histogram(bins = 50, aes(fill = ..count..), color = "black") +
  scale_fill_viridis_c() +
  theme_minimal() +
  ggtitle("Counts per spot")

ggplot(PSC011_all@meta.data, aes(x = nFeature_Spatial)) +
  geom_histogram(bins = 50, aes(fill = ..count..), color = "#080912") +
  scale_fill_gradient(low = "#CDE5E9", high = "#4EA2B1") +
  theme_light() +
  ggtitle("Features per spot")

ggplot(PSC011_all@meta.data, aes(x = pxl_col, y = pxl_row, color = factor(in_tissue))) +
  geom_point(size = 3.0, alpha = 0.9) +
  scale_color_manual(values = c("0" = "#7C825B", "1" = "#E0FF05")) +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  ggtitle("In Tissue vs Background Spots")

ggplot(PSC011_A1@meta.data, aes(x = pxl_col, y = pxl_row, color = nFeature_Spatial)) + # REPLACE WITH nFeature_Spatial for feaures per spot
  annotation_custom(PSC011_A1_img, #replace with PSC011_A1_img, PSC011_B1_img, PSC011_C1_img, and PSC011_D1_img, respectively
                    xmin = 0,
                    xmax = max(PSC011_all@meta.data$pxl_col),
                    ymin = 0,
                    ymax = max(PSC011_all@meta.data$pxl_row)) +
  geom_point(size = 3.0, alpha = 0.9, shape = 16) +
  scale_color_viridis_c() +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  ggtitle("H&E imaging PSC tissue: Features per Spot: A1")

# At this point, I have picked the one image that best match to PSC011_all, that is PSC011_A1_img
# Therefore, I decide to Work on a single replicate because I only have one biological tissue, repeated 4 times
# The replicates are technical repeats, not independent biological samples
# Integration algorithms are designed to align independent biological samples
# Rare cell populations (Kupffer, plasma cells, cholangiocytes) may be underrepresented in technical replicates

```



```{r}
#-----------------------------------------------
# Step 5: Visualize feature variables (PSC01_A1)
#------------------------------------------------
DefaultAssay(PSC011_A1) <- "Spatial"
PSC011_A1 <- NormalizeData(PSC011_A1)                                                                                                             
PSC011_A1 <- FindVariableFeatures(PSC011_A1)
PSC011_A1 <- ScaleData(PSC011_A1)
PSC011_A1 <- RunPCA(PSC011_A1)
ElbowPlot(PSC011_A1)
PSC011_A1 <- FindNeighbors(PSC011_A1, dims = 1:20)
PSC011_A1 <- FindClusters(PSC011_A1, resolution = 0.5)
PSC011_A1 <- RunUMAP(PSC011_A1, dims = 1:20)

DimPlot(PSC011_A1, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```


```{r}
markers_A1 <- FindAllMarkers(PSC011_A1)
head(markers_A1)
heatmap_data <- markers_A1 %>%
  select(cluster, gene, avg_log2FC) %>%
  pivot_wider(names_from = cluster, values_from = avg_log2FC)
# Find 30 markers for all clusters
heatmap_long <- heatmap_data %>%
  pivot_longer(cols = -gene, names_to = "cluster", values_to = "avg_log2FC")
top50 <- heatmap_long %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 50) %>%
  ungroup()
top50 <- top50 %>%
  mutate(gene = str_replace(gene, "\\.\\d+$", ""))
top50_genes <- unique(top50$gene)
DoHeatmap(PSC011_A1, features = top50_genes) +
  scale_fill_gradientn(colors = brewer.pal(n = 15, name = "YlGnBu")) +
  theme(legend.position = "right")

# At this point, we found out that there are 14,866 genes among total of 7 clusters (0 -> 6) within replicate PSC01_A1
# Now, we need to find out cell types for each cluster in PSC011_A1.
# This is a liver tissue with a disorder of Primary Sclerosing cholangitis, therefore average proportions may differ from the healthy adult liver tissue
```



```{r}
#---------------------------
# step 6: Assign cell types  
#---------------------------
rownames(PSC011_A1) <- str_replace(rownames(PSC011_A1), "\\.\\d+$", "")
DotPlot(
  PSC011_A1, 
  group.by = "seurat_clusters",
  features = c('ALB', 'TTR', 'APOA1', 'APOA2','FGB', 'FGG', 'CYP3A4', 'HAMP', "GLUL", "CPS1", "HAL", "ASS1", 
               'KRT19', 'KRT7', 'EPCAM', 'SOX9', 'ANXA4', 'CLDN4',
               'CD68', 'MARCO', 'VSIG4', 'CD163', 'CSF1R', 'C1QA',
               'PECAM1', 'VWF', 'CD34', 'LYVE1', 'SELE', 'STAB1','CLEC4G',
               'ACTA2', 'COL1A1', 'COL3A1', 'PDGFRB', 'VIM', 'DES',
               'CD3D', 'CD3E', 'TRAC', 'CD4', 'CD8A', 'CD8B',
               'MS4A1', 'CD79A', 'CD79B', 'IGKC','JCHAIN', 'IGHM', 'IGHG1'),
  ) + RotatedAxis()

FeaturePlot(PSC011_A1, features = c("CD68","MARCO","CD163","VSIG4","C1QA")) # Kufpffer cell
FeaturePlot(PSC011_A1, features = c("JCHAIN","CD79A","CD79B","IGKC")) # Plasma markers

# Annotate clusters with known liver cell types
PSC011_A1$cell_types <- plyr::mapvalues(PSC011_A1$seurat_clusters,
                                   from = c("0","1","2","3","4","5",'6'),
                                   to   = c("Hepatocyte 1", 
                                            "Hepatocyte 2",
                                            "Plasma cells",
                                            "Cholangiocytes",
                                            "Stellate cells",
                                            "Endothelial cells",
                                            "Kupffer cells"))
DimPlot(
  PSC011_A1,
  group.by = "cell_types",
  label = TRUE,
  repel = TRUE) + theme_classic()
```



```{r}
#---------------------------
# step 7: Spatial coordinates: checking scale factors_json.json  
#---------------------------

table(PSC011_A1@meta.data$in_tissue)

# spot_diameter_fullres": 214.05155323534254, "tissue_hires_scalef": 0.0770416,
#"fiducial_diameter_fullres": 345.7755859955533, "tissue_lowres_scalef": 0.023112481}
# 214.05 × 0.0770416 ≈ ~16.5 PIXELS

scale_factor <- 0.0770416 #tissue_hires_scalef 
df1<- data.frame(
  x = PSC011_A1@meta.data$pxl_col*scale_factor,
  y = PSC011_A1@meta.data$pxl_row*scale_factor,
  cell_type = PSC011_A1@meta.data$cell_types,
  in_tissue = PSC011_A1@meta.data$in_tissue)

df1<- df1[df1$in_tissue == 1, ]

ggplot() +
  annotation_raster(
    PSC011_A1_img$raster,
    xmin = min(df1$x),
    xmax = max(df1$x),
    ymin = min(df1$y),
    ymax = max(df1$y)
  ) +
  geom_point(
    data = df1,
    aes(x = x, y = y, fill = cell_type),
    shape = 21,
    size = 3.0,
    stroke = 0
  ) +
  scale_fill_brewer(palette = "Set2", name = "Cell types") +
  scale_y_reverse() +
  coord_fixed() +
  theme_void()

# Spatial gene expression plot
gene <- "IGKC"
df_expr <- data.frame(
  x = PSC011_A1@meta.data$pxl_col*scale_factor,
  y = PSC011_A1@meta.data$pxl_row*scale_factor,
  expression = FetchData(PSC011_A1, vars = gene)[,1],
  in_tissue = PSC011_A1@meta.data$in_tissue
)

df_expr <- df_expr[df_expr$in_tissue == 1, ]

ggplot() +
  annotation_raster(
    PSC011_A1_img$raster,
    xmin = min(df_expr$x),
    xmax = max(df_expr$x),
    ymin = min(df_expr$y),
    ymax = max(df_expr$y)
  ) +
  geom_point(
    data = df_expr,
    aes(x = x, y = y, color = expression),
    size = 3.0
  ) +
  scale_color_gradientn(
    colors = c("blue","green","yellow","red"),
    name = gene
  ) +
  scale_y_reverse() +
  coord_fixed() +
  theme_void()
```


```{r}
#---------------------------
# step 8: Fibrosis Septa 
#---------------------------

fibrosis_genes <- c(
  "ACTA2",   # myofibroblast marker recommended by GPT
  "COL1A1",
  "COL1A2",
  "COL3A1",
  "COL5A1",
  "PDGFRB",
  "TIMP1",
  "MMP2",
  "SPARC",
  "TAGLN")

cholangiocyte_genes <- c(
  "KRT19",           ## cholangiocytes markers recommended by GPT
  "KRT7",
  "EPCAM",
  "SOX9",
  "MUC1")

PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(fibrosis_genes),
  name = "Fibrosis_Score")
PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(cholangiocyte_genes),
  name = "Cho_Score")
PSC011_A1$portal_fibrosis_score <- PSC011_A1$Fibrosis_Score1 + PSC011_A1$Cho_Score1

df_zone <- data.frame(
  x = PSC011_A1$pxl_col * scale_factor,
  y = PSC011_A1$pxl_row * scale_factor,
  zonation = PSC011_A1$portal_fibrosis_score,
  in_tissue = PSC011_A1$in_tissue
)

df_zone <- df_zone[df_zone$in_tissue == 1, ]

ggplot() +
  annotation_raster(
    PSC011_A1_img$raster,
    xmin=min(df_zone$x),
    xmax=max(df_zone$x),
    ymin=min(df_zone$y),
    ymax=max(df_zone$y)
  ) +
  geom_point(
    data=df_zone,
    aes(x=x, y=y, color=zonation),
    size=3.0
  ) +
  scale_color_gradient2(
    low="blue",
    mid="#800080",
    high="#FFFF00",
    midpoint=0,
    name="Fibrosis Septa"
  ) +
  scale_y_reverse() +
  coord_fixed() +
  theme_void()

df_plot <- data.frame(
  cell_type = PSC011_A1@meta.data$cell_types,
  zonation = PSC011_A1$portal_fibrosis_score)
df_plot <- df_plot[PSC011_A1$in_tissue == 1, ]
df_summary <- df_plot %>%
  group_by(cell_type) %>%
  summarise(mean_zonation = mean(zonation))
ggplot(df_plot, aes(x = cell_type, y = zonation, fill = cell_type)) +
  stat_summary(fun = mean, geom = "bar", color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(y = "Fibrosis Septa") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none")
```



```{r}
#---------------------------
# step 9: Zonation Score: Portal vs Central Zonation
#---------------------------

central_genes <- c(
  "GLUL",   # glutamine synthetase (classic CV marker)
  "CYP2E1",
  "CYP1A2",
  "CYP3A4",
  "LGR5",
  "OAT")
portal_genes <- c(
  "CPS1",
  "HAL",
  "ASS1",
  "ALB",
  "HAMP",
  "PCK1")

PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(central_genes),
  name = "CentralScore")
PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(portal_genes),
  name = "PortalScore")

PSC011_A1$zonation_score  <-  PSC011_A1$CentralScore1 - PSC011_A1$PortalScore1

df_zone1<- data.frame(
  x = PSC011_A1$pxl_col * scale_factor,
  y = PSC011_A1$pxl_row * scale_factor,
  zonation = PSC011_A1$zonation_score,
  in_tissue = PSC011_A1$in_tissue
)

df_zone1 <- df_zone1[df_zone1$in_tissue == 1, ]

ggplot() +
  annotation_raster(
    PSC011_A1_img$raster,
    xmin=min(df_zone1$x),
    xmax=max(df_zone1$x),
    ymin=min(df_zone1$y),
    ymax=max(df_zone1$y)
  ) +
  geom_point(
    data=df_zone1,
    aes(x=x, y=y, color=zonation),
    size=2.5
  ) +
  scale_color_gradient2(
    low="blue",
    mid="#FFFF00",
    high="#800080",
    midpoint=0,
    name="Zonation score"
  ) +
  scale_y_reverse() +
  coord_fixed() +
  theme_void()

df_plot1<- data.frame(
  cell_type = PSC011_A1@meta.data$cell_types,
  zonation = PSC011_A1$zonation_score)
df_plot1 <- df_plot1[PSC011_A1$in_tissue == 1, ]
df_summary1<- df_plot1%>%
  group_by(cell_type) %>%
  summarise(mean_zonation = mean(zonation))
ggplot(df_plot1, aes(x = cell_type, y = zonation, fill = cell_type)) +
  stat_summary(fun = mean, geom = "bar", color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(y = "Zonation score") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none")
hist(PSC011_A1$CentralScore1)
hist(PSC011_A1$PortalScore1)
```



```{r}
#--------------------------------------------------------
# step 9: Tumor-stroma interfaces
#--------------------------------------------------------

tumor_genes <- c(
  "EPCAM",
  "KRT8",
  "KRT18",
  "KRT19",
  "MUC1")
stroma_genes <- c(
  "COL1A1",
  "COL3A1",
  "ACTA2",
  "SPARC",
  "DCN",
  "TAGLN")

PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(tumor_genes),
  name = "TumorScore")
PSC011_A1 <- AddModuleScore(
  PSC011_A1,
  features = list(stroma_genes),
  name = "StromaScore")
PSC011_A1$interface_score <- 
  PSC011_A1$TumorScore1 + PSC011_A1$StromaScore1

df_interface <- data.frame(
  x = PSC011_A1$pxl_col * scale_factor,
  y = PSC011_A1$pxl_row * scale_factor,
  interface = PSC011_A1$interface_score,
  in_tissue = PSC011_A1$in_tissue)
df_interface <- df_interface[df_interface$in_tissue == 1, ]

ggplot() +
  annotation_raster(
    PSC011_A1_img$raster,
    xmin=min(df_interface$x),
    xmax=max(df_interface$x),
    ymin=min(df_interface$y),
    ymax=max(df_interface$y)
  ) +
  geom_point(
    data=df_interface,
    aes(x=x, y=y, color=interface),
    size=3
  ) +
  scale_color_gradient2(
    low="blue",
    mid="purple",
    high="yellow",
    midpoint=0,
    name="Tumor–Stroma Interface"
  ) +
  scale_y_reverse() +
  coord_fixed() +
  theme_void()

df_plot2 <- data.frame(
  cell_type = PSC011_A1@meta.data$cell_types,
  interface = PSC011_A1$interface_score)

ggplot(df_plot2, aes(x = cell_type, y = interface, fill = cell_type)) +
  stat_summary(fun = mean, geom = "bar", color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(y = "Tumor–Stroma Interface") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none")
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
